---
title: 'Foundations for statistical inference - Sampling distributions'
author: "Md Forhad Akbar"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

```{r load-package, eval=TRUE}
library(e1071)
```

In this lab, we investigate the ways in which the statistics from a random 
sample of data can serve as point estimates for population parameters.  We're 
interested in formulating a *sampling distribution* of our estimate in order 
to learn about the properties of the estimate, such as its distribution.

## The data

We consider real estate data from the city of Ames, Iowa.  The details of 
every real estate transaction in Ames is recorded by the City Assessor's 
office.  Our particular focus for this lab will be all residential home sales 
in Ames between 2006 and 2010.  This collection represents our population of 
interest.  In this lab we would like to learn about these home sales by taking 
smaller samples from the full population.  Let's load the data.

```{r load-data, eval=TRUE}
load("more/ames.RData")
```

We see that there are quite a few variables in the data set, enough to do a 
very in-depth analysis.  For this lab, we'll restrict our attention to just 
two of the variables: the above ground living area of the house in square feet 
(`Gr.Liv.Area`) and the sale price (`SalePrice`).  To save some effort 
throughout the lab, create two variables with short names that represent these 
two variables.  

```{r assign, eval=TRUE}
area <- ames$Gr.Liv.Area
price <- ames$SalePrice
```

Let's look at the distribution of area in our population of home sales by 
calculating a few summary statistics and making a histogram.

```{r area, eval=TRUE}
summary(area)
hist(area)
```

1.  Describe this population distribution.  
**Answer:**

```{r}
IQR(area)
area_sd<-sd(area)
area_sd
area_mean<-mean(area)
area_mean
```

```{r qq-area}
qqnorm(area)
qqline(area, col='blue')
```

```{r histogram-area}
hist(area, probability = TRUE, col = 'pink')
x <- 0:6000
y <- dnorm(x = x, mean = area_mean, sd = area_sd)
lines(x = x, y = y, col = "blue")
```

```{r skewness-area}
# use skewness function from e1071
skewness(area)
```

We can conclude from above that the living area distribution of Ames is a right skewed distribution, e.g. most of the living areas are small: Half of the houses have less than 1450 sq.feet, but there are also some houses with more than 4000 square feet living space. The middle 50 percent of the homes have areas between 1126 and 1742 sq.ft.  

## The unknown sampling distribution

In this lab we have access to the entire population, but this is rarely the 
case in real life.  Gathering information on an entire population is often 
extremely costly or impossible.  Because of this, we often take a sample of 
the population and use that to understand the properties of the population.

If we were interested in estimating the mean living area in Ames based on a 
sample, we can use the following command to survey the population.

```{r samp1, eval=TRUE}
samp1 <- sample(area, 50)
```

This command collects a simple random sample of size 50 from the vector 
`area`, which is assigned to `samp1`.  This is like going into the City 
Assessor's database and pulling up the files on 50 random home sales.  Working 
with these 50 files would be considerably simpler than working with all 2930 
home sales.

2.  Describe the distribution of this sample. How does it compare to the 
    distribution of the population?  
**Answer:**

```{r samp1-summary, eval=TRUE}
summary(samp1)
hist(samp1)
```

```{r}
IQR(samp1)
samp1_sd<-sd(samp1)
samp1_sd
samp1_mean<-mean(samp1)
samp1_mean
```

```{r qq-samp1}
qqnorm(samp1)
qqline(samp1, col='blue')
```

 
```{r histogram-samp1}
hist(samp1, probability = TRUE, col = 'pink')
x <- 0:2500
y <- dnorm(x = x, mean = samp1_mean, sd = samp1_sd)
lines(x = x, y = y, col = "blue")
```   
    
```{r skewness-samp1}
# use skewness function from e1071
skewness(samp1)
```   
    
We can conclude from above that Samp1 distribution is also right skewed but skewness for sam1 (0.6583092) is much less compare to population (1.272805) distribution.  
    
If we're interested in estimating the average living area in homes in Ames 
using the sample, our best single guess is the sample mean.

```{r mean-samp1, eval=TRUE}
mean(samp1)
```

Depending on which 50 homes you selected, your estimate could be a bit above 
or a bit below the true population mean of 1499.69 square feet.  In general, 
though, the sample mean turns out to be a pretty good estimate of the average 
living area, and we were able to get it by sampling less than 3\% of the 
population.



3.  Take a second sample, also of size 50, and call it `samp2`.  How does the 
    mean of `samp2` compare with the mean of `samp1`?  Suppose we took two 
    more samples, one of size 100 and one of size 1000. Which would you think 
    would provide a more accurate estimate of the population mean?  
**Answer:**

```{r samp2, eval=TRUE}
samp2 <- sample(area, 50)
```

```{r samp2-summary, eval=TRUE}
summary(samp2)
```

```{r samp3, eval=TRUE}
samp3 <- sample(area, 100)
```

```{r samp3-summary, eval=TRUE}
summary(samp3)
```

```{r samp4, eval=TRUE}
samp4 <- sample(area, 1000)
```

```{r samp4-summary, eval=TRUE}
summary(samp4)
```

We can see from above that mean of samp1 is 1451.44 when samp2 has a higher mean 1588. I think mean of 1000 sample would be more accurate estimate of the population mean. We can see from above thta population mean is 1499.69 while samp3 (sample size 100) mean is 1581 and samp4 (sample size 1000) mean is 1485. So, we can conclude the mean of 1000 sample is more accurate estimate of the population mean.  

Not surprisingly, every time we take another random sample, we get a different 
sample mean.  It's useful to get a sense of just how much variability we 
should expect when estimating the population mean this way. The distribution 
of sample means, called the *sampling distribution*, can help us understand 
this variability. In this lab, because we have access to the population, we 
can build up the sampling distribution for the sample mean by repeating the 
above steps many times. Here we will generate 5000 samples and compute the 
sample mean of each.

```{r loop, eval=TRUE}
sample_means50 <- rep(NA, 5000)

for(i in 1:5000){
   samp <- sample(area, 50)
   sample_means50[i] <- mean(samp)
   }

hist(sample_means50)
```

If you would like to adjust the bin width of your histogram to show a little 
more detail, you can do so by changing the `breaks` argument.

```{r hist-breaks, eval=TRUE}
hist(sample_means50, breaks = 25)
```

Here we use R to take 5000 samples of size 50 from the population, calculate 
the mean of each sample, and store each result in a vector called 
`sample_means50`. On the next page, we'll review how this set of code works.

4.  How many elements are there in `sample_means50`?  Describe the sampling 
    distribution, and be sure to specifically note its center.  Would you 
    expect the distribution to change if we instead collected 50,000 sample 
    means?  

**Answer:**

```{r}
str(sample_means50)
```
    
```{r summary-sample_means50}
summary(sample_means50)
```  

```{r loop-sample_means500, eval=TRUE}
sample_means500 <- rep(NA, 50000)

for(i in 1:50000){
   samp <- sample(area, 50)
   sample_means500[i] <- mean(samp)
   }

hist(sample_means500)
```
```{r summary-sample_means500}
summary(sample_means500)
```
There are 5000 elements in sample_means50. We can see that the distribution of 50000 sample looks pretty much same as distribution of mean of 5000. Both has the same mean 1500 and meadian 1498.  
    

## Interlude: The `for` loop

Let's take a break from the statistics for a moment to let that last block of 
code sink in.  You have just run your first `for` loop, a cornerstone of 
computer programming.  The idea behind the for loop is *iteration*: it allows 
you to execute code as many times as you want without having to type out every 
iteration.  In the case above, we wanted to iterate the two lines of code 
inside the curly braces that take a random sample of size 50 from `area` then 
save the mean of that sample into the `sample_means50` vector.  Without the 
`for` loop, this would be painful:

```{r loop-long, eval=TRUE}
sample_means50 <- rep(NA, 5000)

samp <- sample(area, 50)
sample_means50[1] <- mean(samp)

samp <- sample(area, 50)
sample_means50[2] <- mean(samp)

samp <- sample(area, 50)
sample_means50[3] <- mean(samp)

samp <- sample(area, 50)
sample_means50[4] <- mean(samp)
```

and so on...

With the for loop, these thousands of lines of code are compressed into a 
handful of lines. We've added one extra line to the code below, which prints 
the variable `i` during each iteration of the `for` loop. Run this code.

```{r loop-again, eval=FALSE}
sample_means50 <- rep(NA, 5000)

for(i in 1:5000){
   samp <- sample(area, 50)
   sample_means50[i] <- mean(samp)
   print(i)
   }
```

Let's consider this code line by line to figure out what it does.  In the 
first line we *initialized a vector*.  In this case, we created a vector of 
5000 zeros called `sample_means50`.  This vector will will store values 
generated within the `for` loop.

The second line calls the `for` loop itself.  The syntax can be loosely read as, 
"for every element `i` from 1 to 5000, run the following lines of code". You 
can think of `i` as the counter that keeps track of which loop you're on. 
Therefore, more precisely, the loop will run once when `i = 1`, then once when 
`i = 2`, and so on up to `i = 5000`.

The body of the `for` loop is the part inside the curly braces, and this set of 
code is run for each value of `i`.  Here, on every loop, we take a random 
sample of size 50 from `area`, take its mean, and store it as the 
$i$<sup>th</sup> element of `sample_means50`.

In order to display that this is really happening, we asked R to print `i` at 
each iteration. This line of code is optional and is only used for displaying 
what's going on while the `for` loop is running.

The `for` loop allows us to not just run the code 5000 times, but to neatly 
package the results, element by element, into the empty vector that we 
initialized at the outset. 

5.  To make sure you understand what you've done in this loop, try running a 
    smaller version.  Initialize a vector of 100 zeros called 
    `sample_means_small`.  Run a loop that takes a sample of size 50 from 
    `area` and stores the sample mean in `sample_means_small`, but only 
    iterate from 1 to 100.  Print the output to your screen (type 
    `sample_means_small` into the console and press enter).  How many elements 
    are there in this object called `sample_means_small`? What does each 
    element represent?  
**Answer:**

```{r loop-sample_means_small, eval=TRUE}
sample_means_small <- rep(NA, 100)

for(i in 1:100){
   samp <- sample(area, 50)
   sample_means_small[i] <- mean(samp)
}
sample_means_small
```
There are 100 elements in the object and each of them represents a sample mean.  

## Sample size and the sampling distribution

Mechanics aside, let's return to the reason we used a `for` loop: to compute a 
sampling distribution, specifically, this one.

```{r hist, eval=TRUE}
hist(sample_means50)
```

The sampling distribution that we computed tells us much about estimating 
the average living area in homes in Ames.  Because the sample mean is an 
unbiased estimator, the sampling distribution is centered at the true average 
living area of the the population, and the spread of the distribution 
indicates how much variability is induced by sampling only 50 home sales.

To get a sense of the effect that sample size has on our distribution, let's 
build up two more sampling distributions: one based on a sample size of 10 and 
another based on a sample size of 100.

```{r samp-10-100, eval=TRUE}
sample_means10 <- rep(NA, 5000)
sample_means100 <- rep(NA, 5000)

for(i in 1:5000){
  samp <- sample(area, 10)
  sample_means10[i] <- mean(samp)
  samp <- sample(area, 100)
  sample_means100[i] <- mean(samp)
}
```

Here we're able to use a single `for` loop to build two distributions by adding 
additional lines inside the curly braces.  Don't worry about the fact that 
`samp` is used for the name of two different objects.  In the second command 
of the `for` loop, the mean of `samp` is saved to the relevant place in the 
vector `sample_means10`.  With the mean saved, we're now free to overwrite the 
object `samp` with a new sample, this time of size 100.  In general, anytime 
you create an object using a name that is already in use, the old object will 
get replaced with the new one.

To see the effect that different sample sizes have on the sampling 
distribution, plot the three distributions on top of one another.

```{r plot-samps, eval=TRUE, tidy = FALSE}
par(mfrow = c(3, 1))

xlimits <- range(sample_means10)

hist(sample_means10, breaks = 20, xlim = xlimits)
hist(sample_means50, breaks = 20, xlim = xlimits)
hist(sample_means100, breaks = 20, xlim = xlimits)
```

The first command specifies that you'd like to divide the plotting area into 3 
rows and 1 column of plots (to return to the default setting of plotting one 
at a time, use `par(mfrow = c(1, 1))`). The `breaks` argument specifies the 
number of bins used in constructing the histogram.  The `xlim` argument 
specifies the range of the x-axis of the histogram, and by setting it equal 
to `xlimits` for each histogram, we ensure that all three histograms will be 
plotted with the same limits on the x-axis.

6.  When the sample size is larger, what happens to the center?  What about the spread?  
**Answer:**
When the sample size is larger the center will be closer to population mean. The spread of the distribution gets smaller as sample size increases

* * *
## On your own

So far, we have only focused on estimating the mean living area in homes in 
Ames.  Now you'll try to estimate the mean home price.

-   Take a random sample of size 50 from `price`. Using this sample, what is 
    your best point estimate of the population mean?  
**Answer:**

```{r samp1-price, eval=TRUE}
price_samp1 <- sample(price, 50)
```

```{r samp1-price-summary, eval=TRUE}
summary(price_samp1)
```
```{r price-summary, eval=TRUE}
summary(price)
```
My best point estimate of the mean sale price of homes in Ames is around 196502, versus the population mean 180796.  

-   Since you have access to the population, simulate the sampling 
    distribution for $\bar{x}_{price}$ by taking 5000 samples from the 
    population of size 50 and computing 5000 sample means.  Store these means 
    in a vector called `sample_means50`. Plot the data, then describe the 
    shape of this sampling distribution. Based on this sampling distribution, 
    what would you guess the mean home price of the population to be? Finally, 
    calculate and report the population mean.  
**Answer:**

```{r}
sample_means50 <- rep(NA, 5000)

for(i in 1:5000){
  sample_means50[i] <- mean(sample(price, 50))
  }

hist(sample_means50, col = "blue", border = "black", breaks = 25, xlim=c(min(sample_means50), max(sample_means50)))
```

```{r}
summary(sample_means50)
```

This sampling distribution is unimodal and very close to normal.
I would estimate of the mean sale price of homes in Ames is around 180947 (mean of sample_means50), which is very close to the population mean 180796.  

-   Change your sample size from 50 to 150, then compute the sampling 
    distribution using the same method as above, and store these means in a 
    new vector called `sample_means150`. Describe the shape of this sampling 
    distribution, and compare it to the sampling distribution for a sample 
    size of 50.  Based on this sampling distribution, what would you guess to 
    be the mean sale price of homes in Ames?  
**Answer:**
```{r}
sample_means150 <- rep(NA, 5000)

for(i in 1:5000){
  sample_means150[i] <- mean(sample(price, 150))
  }
hist(sample_means150, col = "pink", border = "black", breaks = 25, xlim=c(min(sample_means50), max(sample_means50)))
```

```{r}
summary(sample_means150)
```
This sampling distribution is also unimodal and much closer to normal. It spreads less than the distribution for sample size 50. I would guess the mean sale price of homes in Ames is around 180851 which is the mean of sample_means150.  

-   Of the sampling distributions from 2 and 3, which has a smaller spread?  If
    we're concerned with making estimates that are more often close to the 
    true value, would we prefer a distribution with a large or small spread?  
**Answer:**
The sampling distribution for a sample size 150 has a smaller spread than the one for sample size 50. We would prefer a smaller spread distribution, which usually with a larger than sample size (n), to have a point estimate to be closer to the true mean (at most of the time).  
